---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Final Project: Video Game Sales

#### Analyn Seeman, Devon Katragadda, Cassidy Berghoff, and Sofi Gutierrez

## Introduction

Intro goes here -- State the goal of the project

1.  Which genre has the best sales?

2.  Is there a relationship between user rating and sales? What about
    critic rating in sales? Which one would be better marker for a
    bestseller?

3.  Which platform had the most successful games?

4.  Which platform's games have the best rating? Does platform affect
    the above questions?

5.  What is the most popular genre per platform?

... and so on

### Uploading Data Set

```{r}

data = read.csv("Video Game Sales 1978 - 2024 export 2025-03-31 22-10-49.csv")

head(data)


```

### Description of Data Set

```{r}

str(data)

```

This data set compiles data from video game sales from 1978 to 2024.
There are 16 columns in this dataset. They are the following: Rank: This
is the index of the dataset. Name: The name of the game. Platform: The
platform that the game was released for. All_Platforms: A variable that
holds a "list" that describes which platforms a game is on if it is on
multiple. All_Games: A list of all of the games in that series.
Publisher: The publisher of the games. Developer: The people that
developed the game. Critic_Score: What numeric value the critics gave
the game (NA, 0-10) User_Score: What numeric value the users gave the
game (NA, 0-10) NA_Sales: PAL_Sales: JP_Sales: Other_Sales:
Global_Sales: Year: The year the game was released (if in a series, the
year the most recent game was released) Genre: The genre of the game.

### Cleaning of Data Set

```{r}
library(dplyr)
library(tidyr)

q1_data <- data %>%
  drop_na(Global_Sales) #dropping games with no available Sale data

q2_data <- q1_data %>%
  drop_na(Critic_Score) %>%
  drop_na(User_Score) #dropping games with no available Sale data

```

## Marginal Summaries

```{r}

q1_data %>%
  group_by(Platform) %>%
  summarise(maxSales = max(Global_Sales)) #3

# Anything with critic ratings, user ratings, and sales data 
q2_data %>%
  group_by(Platform) %>%
  summarise(maxUserScore = max(User_Score), maxCriticScore = max(Critic_Score)) #4

```

## Conclusion
