---
title: "Final Project: "
author: "Analyn Seeman, Devon Katragadda, Cassidy Berghoff, and Sofi Gutierrez"
date: "3/31/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Uploading Data Set

```{r}

data = read.csv("Video Game Sales 1978 - 2024 export 2025-03-31 22-10-49.csv")

head(data)


```

## Description of Data Set
``` {r}

str(data)

```

This data set compiles data from video game sales from 1978 to 2024. There are 16 columns in this dataset. They are the following:
Rank: This is the index of the dataset.
Name: The name of the game.
Platform: The platform that the game was released for.
All_Platforms: A variable that holds a "list" that describes which platforms a game is on if it is on multiple.
All_Games: A list of all of the games in that series. 
Publisher: The publisher of the games.
Developer: The people that developed the game.
Critic_Score: What numeric value the critics gave the game (NA, 0-10)
User_Score: What numeric value the users gave the game (NA, 0-10)
NA_Sales: 
PAL_Sales: 
JP_Sales:
Other_Sales:
Global_Sales:
Year: The year the game was released (if in a series, the year the most recent game was released)
Genre: The genre of the game.


## Cleaning of Data Set

``` {r}
library(dplyr)
library(tidyr)

data <- data %>%
  drop_na(Global_Sales) #dropping games with no available Sale data

```


## Marginal Summaries


## Questions to be Addressed

Which genre has the best sales? 

Is there a relationship between user rating and sales? What about critic rating in sales? Which one would be better marker for a bestseller?

Which platform had the most successful games?

Which platform's games have the best rating? Does platform affect the above questions?

What is the most popular genre per platform?
