---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Final Project: Video Game Sales

#### Analyn Seeman, Devon Katragadda, Cassidy Berghoff, and Sofi Gutierrez

## Introduction

The video game industry is one of the largest and fastest-growing
entertainment sectors in the world, engaging audiences across the world.
With the rise of digital distribution platforms, online reviews, and
global gaming communities, understanding what contributes to a video
game's commercial and critical success has become more relevant than
ever for developers, publishers, and marketers alike.

This project aims to explore the factors that define a successful video
game by analyzing historical data across multiple categories such as
sales, ratings, platforms, genres, and geographic trends. By examining
these attributes, we hope to identify patterns and correlations that can
offer insights into consumer preferences and industry dynamics.

To guide our analysis, we focus on the following research questions:

1.  Which genre has the best sales?

2.  Is there a relationship between user rating and sales? What about
    critic rating in sales? Which one would be better marker for a
    bestseller?

3.  Which platform had the most successful games?

4.  Which platform's games have the best rating? Does platform affect
    the above questions?

5.  What is the most popular genre per platform?

6.  What years were most successful for video games

7.  During the best years, which genres were most popular?

8.  What is the most popular genre by area?

By answering these questions, we aim to identify key trends that
influence the success of a video game. The insights gained from this
analysis can help video game developers find the best chance of success
in the gaming industry.

## Data

### Uploading Data Set

```{r}

data = read.csv("Video Game Sales 1978 - 2024 export 2025-03-31 22-10-49.csv")

head(data)
```
### Adding the libraries needed
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

### Description of Data Set

```{r}
str(data)
```

This data set compiles data from video game sales from 1978 to 2024.
There are 16 columns in this data set, representing the 16 variables in
the data set.

### Variables

-   Rank: The index of the data set
-   Name: The name of the game
-   Platform: The platform that the game was released for
-   All_Platforms: A variable that holds a "list" that describes which
    platforms a game is on if it is on multiple
-   All_Games: A list of all of the games in that series
-   Publisher: The publisher of the games
-   Developer: The people that developed the game
-   Critic_Score: What numeric value the critics gave the game (NA,
    0-10)
-   User_Score: What numeric value the users gave the game (NA, 0-10)
-   NA_Sales: Sales in the North American region
-   PAL_Sales: Sales in the European region
-   JP_Sales: Sales in the Eastern Asia region
-   Other_Sales: Sales in all other regions
-   Global_Sales: Year: The year the game was released (if in a series,
    the year the most recent game was released)
-   Genre: The genre of the game

### Cleaning of Data Set

```{r}

q1_data <- data %>%
  drop_na(Global_Sales) #dropping games with no available Sale data

q2_data <- q1_data %>%
  drop_na(Critic_Score) %>%
  drop_na(User_Score)   #dropping games with no available Sale data

```

### Marginal Summaries

```{r}
q1_data %>%
  group_by(Platform) %>%
  summarise(maxSales = max(Global_Sales)) #3

# Anything with critic ratings, user ratings, and sales data 
q2_data %>%
  group_by(Platform) %>%
  summarise(maxUserScore = max(User_Score), maxCriticScore = max(Critic_Score))
```

## Results

### Going into Genres

```{r}
q1_genre <- q1_data %>%
  group_by(Genre) %>%
  summarise(totalSales = sum(Global_Sales)) %>%
  arrange(desc(totalSales))

q1_genre
```
#### Which Genre is the Most Popular?
```{r}
highlight_colors <- c(
  "yes"  = "#de7e5d",  # Pastel red
  "no"      = "#edb48c"  # Pastel blue
)

q1_data %>% mutate(highlight = if_else(Genre %in% c("Sports", "Action", "Shooter"), "yes", "no")) %>% ggplot( aes(x = Genre, weight= Global_Sales, fill=highlight)) +
  geom_bar(show.legend = FALSE) +
  scale_fill_manual(values = highlight_colors) +
  theme(axis.text.x = element_text(angle = 90)) +
  ylab("Total Sales") + 
  ggtitle("Global Sales by Genre")
```


The above table shows the total global sales per video game genre. It is
sorted in decreasing order, meaning the most popular genre is located in
the top row. The genre with the highest total sales indicates the most
popular genre globally. Since Sports has the highest totalSale
value, it is the most popular genre globally, followed by Action and Shooter.

#### Which Genre is the Most Popular in Each Region?
```{r}
genre_reigon_sales <- q1_data %>%
  group_by(Genre) %>%
  summarise(
    NA_Sales = sum(NA_Sales, na.rm = TRUE),
    PAL_Sales = sum(PAL_Sales, na.rm = TRUE),
    JP_Sales = sum(JP_Sales, na.rm = TRUE),
    Other_Sales = sum(Other_Sales, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(NA_Sales, PAL_Sales, JP_Sales, Other_Sales),
    names_to = "Region",
    values_to = "Sales"
  ) %>%
  mutate(
    Region = recode(Region,
                    "NA_Sales" = "North America",
                    "PAL_Sales" = "Europe",
                    "JP_Sales" = "Japan",
                    "Other_Sales" = "Other"),
    highlight = if_else(
      ((Region == "Japan" & (Genre %in% c("Role-Playing", "Platform", "Sports"))) |
         (Region != "Japan" & (Genre %in% c("Sports", "Action", "Shooter")))
         ),"yes", "no" )
  )

highlight_colors <- c(
  "yes"  = "#de7e5d",  # Pastel red
  "no"      = "#edb48c"  # Pastel blue
)


ggplot(genre_reigon_sales, aes(x = Genre, y = Sales, fill = highlight)) +
  geom_col(show.legend = FALSE) +
  theme(axis.text.x = element_text(size = 8, angle = 90, vjust=0.5)) + 
  scale_fill_manual(values = highlight_colors) +
  facet_wrap(~ Region, scales = "free_y") +
  labs(
    title = "Genre Sales by Region",
    x = "Genre",
    y = "Total Sales (in millions)"
  ) 

```

All areas except for Japan have the same Top 3 Genres (Sports, Action, Shooting). Japan however, has Role-Playing, Platforming, and Sports as their top 3. 
Japan is a capital of video gaming. According to IMDb, there exists a genre titled JRPG - or Japanese Role Playing Games. This sub-genre contains popular titles such as Final Fantasy, Kingdom Hearts, and Persona 5. While those titles are popular even in the US, they are obviously extremely popular in Japanese culture given that they are in the sub-genre named after it. 
Source: https://www.imdb.com/list/ls022397153/

#### How does Average Sales affect Popular Genres?
``` {r}
q1_data %>% mutate(highlight = if_else(Genre %in% c("Sandbox", "Party", "Action-Adventure"), "yes", "no")) %>% ggplot( aes(x = Genre, y=Global_Sales, fill=highlight)) +
  geom_bar(stat = "summary", fun = "mean", show.legend=FALSE) +
  scale_fill_manual(values = highlight_colors) +
  theme(axis.text.x = element_text(angle = 90)) +
  ylab("Total Sales (in millions)") + 
  ggtitle("Average Sales by Genre")

```

### Going into User and Critic Scores

#### Do User and Critic Ratings Differ?

```{r}
# Histogram: Distribution of Critic Scores
ggplot(q2_data, aes(x = Critic_Score)) +
  geom_histogram(show.legend = FALSE, binwidth = 1, fill = "#edb48c") +
  labs(title = "Distribution of Critic Score", x = "Critic Score", y = "Count") + 
  scale_x_continuous(breaks = seq(0, 10, by = 1), limits = c(0,11)) + 
  ylim(0,100)

# Histogram: Distribution of User Scores
ggplot(q2_data, aes(x = User_Score)) +
  geom_histogram(show.legend = FALSE, binwidth = 1, fill = "#edb48c") +
  labs(title = "Distribution of User Score", x = "User Score", y = "Count") +
  scale_x_continuous(breaks = seq(0, 10, by = 1), limits = c(0,11)) + 
  ylim(0,100)
```

The distribution of user scores are Left Skewed with an average of 8.54185. While critic scores are less (still left) skewed than user scpres. With an average of 8.162555

```{r}
# Summary Statistics: Mean and Standard Deviation for Scores
q2_data %>%
  summarise(
    avgCS = mean(Critic_Score), 
    sdCs = sd(Critic_Score), 
    avgUS = mean(User_Score), 
    sdUs = sd(User_Score)
  )
```

#### Do Scores Differ Between Users and Critics?
```{r}
# Scatter Plot: Critic Score vs User Score
ggplot(q2_data, aes(x = Critic_Score, y = User_Score)) +
  geom_point() +
  labs(title = "Critic Score vs User Score", x = "Critic Score", y = "User Score")
```

There is definitely a correlation between Critic Score and User Score, but there is a notable difference as the scores get lower.

#### Do Ratings Influence Sales?

```{r}
# Scatter plot for user score vs global sales
ggplot(q2_data, aes(x = User_Score, y = Global_Sales)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "User Score vs Global Sales", x = "User Score", y = "Global Sales")

# Scatter plot for user score vs log global sales
ggplot(q2_data, aes(x = User_Score, y = log(Global_Sales))) +
  geom_point() +
  geom_smooth(method = "lm", col = "#de7e5d") +
  labs(title = "User Score vs Log Global Sales", x = "User Score", y = "Log of Global Sales")
```
```{r}
# Scatter plot for critic score vs global sales
ggplot(q2_data, aes(x = Critic_Score, y = Global_Sales)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Critic Score vs Global Sales", x = "Critic Score", y = "Global Sales")

# Scatter plot for critic score vs log global sales
ggplot(q2_data, aes(x = Critic_Score, y = log(Global_Sales))) +
  geom_point() +
  geom_smooth(method = "lm", col = "#de7e5d") +
  labs(title = "Critic Score vs Log Global Sales", x = "Critic Score", y = "Log of Global Sales")
```

There is a very apparent log linear relationship between Sales and User Score. Critic reviews may hold more weight in shaping purchasing decisions because Less Deviance

#### Do Scores Differ by Platform?
```{r}
q2_platform <- q2_data %>%
  group_by(Platform) %>%
  summarise(
    `Average User Score` = mean(User_Score, na.rm = TRUE),
    `Average Critic Score` = mean(Critic_Score, na.rm = TRUE)
  ) %>%
  arrange(desc(`Average User Score`), desc(`Average Critic Score`))

q2_platform
```

The Scores do differ by platform, but we noticed that there are more apparent patterns between Platform and Sales.

### Going into Platform

#### Which Platform Had the Best Selling Games?

```{r}

# Add Company column to high_sales_platforms
high_sales_platforms <- q1_data %>%
  filter(
         Platform %in% c("N64", "XB", "PS4", "PSP", "PS2", "PS3", "PSN", 
                         "NS", "PC", "X360", "DS", "Wii", "NES", "GC", 
                         "XOne", "PS")) %>%
  mutate(
    Company = case_when(
      Platform %in% c("N64", "NS", "Wii", "DS", "NES", "GC") ~ "Nintendo",
      Platform %in% c("PS", "PS2", "PS3", "PS4", "PSP", "PSN") ~ "Sony",
      Platform %in% c("XB", "X360", "XOne") ~ "Microsoft",
      Platform == "PC" ~ "PC"
    )
  )

# Define custom color mapping
company_colors <- c(
  "Nintendo"  = "#F4A6A6",  # Pastel red
  "Sony"      = "#A6B8E6",  # Pastel blue
  "Microsoft" = "#A9D6A4",  # Pastel green
  "PC"        = "#C0C0C0"   # Light gray
)


# Plot with custom colors
ggplot(high_sales_platforms, aes(x = Platform, y = log(Global_Sales), fill = Company)) +
  geom_boxplot(outlier.color = "red") +
  scale_fill_manual(values = company_colors) +
  labs(
    title = "Distribution of Global Sales by Platform",
    x = "Platform",
    y = "Global Sales (in millions)",
    fill = "Platform Brand"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )

```

We’re looking at the distribution of global sales across gaming platforms. Each box represents a platform, and it shows how game sales are spread out, from lower-performing titles to hits.
If we look at the distribution, we’ll notice that Nintendo platforms like the NES (Nintendo Entertainment System) and Wii show a much wider spread, with some major outliers. The biggest outlier here, over 80 million units, is Wii Sports, which was bundled with the console and became one of the best-selling games of all time.
So overall, Nintendo dominates in iconic, high-performing titles, and Sony and Microsoft stands out in consistent quality.

#### Which Platform is the Best Selling Globally?
```{r}
# Define brand colors
company_colors <- c(
  "Nintendo"  = "#F4A6A6",  # Pastel red
  "Sony"      = "#A6B8E6",  # Pastel blue
  "Microsoft" = "#A9D6A4",  # Pastel green
  "PC"        = "#C0C0C0"   # Light gray
)

# Add Company info to platform summary
q1_platform <- q1_data %>%
  group_by(Platform) %>%
  summarise(totalSales = sum(Global_Sales, na.rm = TRUE)) %>%
  mutate(
    Company = case_when(
      Platform %in% c("N64", "NS", "Wii", "DS", "NES", "GC") ~ "Nintendo",
      Platform %in% c("PS", "PS2", "PS3", "PS4", "PSP", "PSN") ~ "Sony",
      Platform %in% c("XB", "X360", "XOne") ~ "Microsoft",
      Platform == "PC" ~ "PC",
      TRUE ~ "Other"
    )
  ) %>%
  filter(Company != "Other") %>%
  arrange(desc(totalSales))

# Plot with color-coded bars
ggplot(q1_platform, aes(x = reorder(Platform, -totalSales), y = totalSales, fill = Company)) +
  geom_col() +
  scale_fill_manual(values = company_colors) +
  labs(
    title = "Total Global Sales by Platform",
    subtitle = "Color-coded by Platform Brand",
    x = "Platform",
    y = "Total Global Sales (in millions)",
    fill = "Brand"
  ) +
  scale_y_continuous(labels = scales::comma_format()) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    axis.text.y = element_text(size = 10)
  )
```
We’re looking at total global sales across platforms, so instead of individual games, this is the sum of all game sales per platform.
Right away, you can see that the PlayStation 2 is the highest. It’s the most successful platform globally, with over 1.25 billion units sold.
After the PlayStation 2, we see strong performance from XBox 360, the PlayStation 3, Wii, and DS. 
The bars are color-coded by brand, blue for Sony, green for Microsoft, red for Nintendo, and gray for PC, so we can get a sense of which companies had the biggest impact.
It’s also worth noting that while Nintendo had some of the highest-selling individual games, like Wii Sports, Sony had more overall success when you look at total game sales across its platforms.
So overall, this chart really highlights how dominant Sony, especially with the the PlayStation 2

```{r}
# Define company colors
company_colors <- c(
  "Nintendo"  = "#F4A6A6",  # Pastel red
  "Sony"      = "#A6B8E6",  # Pastel blue
  "Microsoft" = "#A9D6A4",  # Pastel green
  "PC"        = "#C0C0C0"   # Light gray
)

# Summarize and reshape data
platform_region_sales <- q1_data %>%
  group_by(Platform) %>%
  summarise(
    NA_Sales = sum(NA_Sales, na.rm = TRUE),
    PAL_Sales = sum(PAL_Sales, na.rm = TRUE),
    JP_Sales = sum(JP_Sales, na.rm = TRUE),
    Other_Sales = sum(Other_Sales, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(NA_Sales, PAL_Sales, JP_Sales, Other_Sales),
    names_to = "Region",
    values_to = "Sales"
  ) %>%
  mutate(
    Region = recode(Region,
                    "NA_Sales" = "North America",
                    "PAL_Sales" = "Europe",
                    "JP_Sales" = "Japan",
                    "Other_Sales" = "Other"),
    Company = case_when(
      Platform %in% c("N64", "NS", "Wii", "DS", "NES", "GC") ~ "Nintendo",
      Platform %in% c("PS", "PS2", "PS3", "PS4", "PSP", "PSN") ~ "Sony",
      Platform %in% c("XB", "X360", "XOne") ~ "Microsoft",
      Platform == "PC" ~ "PC",
      TRUE ~ "Other"
    )
  ) %>%
  filter(Company != "Other")  # remove unknown brands

# Plot: Facet by Region, color by Company
ggplot(platform_region_sales, aes(x = reorder(Platform, -Sales), y = Sales, fill = Company)) +
  geom_col(show.legend = TRUE) +
  scale_fill_manual(values = company_colors) +
  facet_wrap(~ Region, scales = "free_y") +
  labs(
    title = "Platform Sales by Region",
    x = "Platform",
    y = "Total Sales (in millions)",
    fill = "Brand"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10, angle = 60, hjust = 1, vjust = 1),
    strip.placement = "outside",
    strip.background = element_blank(),
    panel.spacing = unit(1.5, "lines"),
    plot.title = element_text(face = "bold", size = 16)
  )
```
This  breaks down platform sales by region, so we can see which consoles were most popular in different parts of the world.
Let’s start with North America and Europe. You’ll notice the pattern is pretty similar in both regions, the PlayStation 2 was popular, followed by Xbox 360, PlayStation 3, and Wii. These were the go-to consoles during the peak gaming years of the 2000s.
In the "Other" category, which consists of the rest of the world excluding Japan, we see the same top performers again: PlayStation 2, Xbox 360, PlayStation 3, and also PlayStation 4.
But we should notice that things look very different in Japan. Here, Nintendo dominates, especially with the DS and NES. The PlayStation 2 is still in the most popular platforms, but Nintendo's presence is much stronger compared to the rest of the world. That really highlights Japan's different gaming culture and preference for handhelds and classic titles.
The chart is once again color-coded by brand, making it easy spot which company led where.
Overall, this comparison shows how Sony had global dominance, but Nintendo remained incredibly strong in Japan, with Microsoft’s biggest impact being in North America and Europe.
This led us into thinking why does Japan stands out so much, not just in our data, but in the history of video games overall.
As Blake Harris puts it in his book Console Wars: “Without the contributions of Japan, we wouldn’t have a video game industry.” That really sets the tone for understanding Japan's influence.
Looking at the data, the top platforms in Japan were the PlayStation 2, DS, standing for Dual Screen, PlayStation, and NES, standing for Nintendo Entertainment System. All of these consoles were created by Sony or Nintendo.
Both of those companies were founded in Japan, with Nintendo headquartered in Kyoto, and Sony based in Tokyo.
And according to CNN, Tokyo was the front leader for gaming, with both Nintendo and Sony having deep roots there. That historical presence helped shape Japan into a gaming powerhouse and explains why platform sales are especially high there.
So when we see Japan’s platform preferences differ from other regions, it’s not random, it reflects deep cultural ties, brand loyalty, and the fact that Japan is a hub for gaming innovation.
https://www.cnn.com/2017/11/12/asia/future-japan-videogame-landmarks/index.html
As said by CNN, Tokyo was the front leader for gaming with SEGA and Nintendo originating there. This explains the platforming being higher. 

### Year Graphs
```{r}
 q1_data %>%
  ggplot(aes(x = Year, y = Global_Sales)) +
  geom_col() +
  labs(
    title = "Global Video Game Sales Over Time",
    x = "Year",
    y = "Global Sales (in millions)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::comma)
```


```{r}
q1_data %>%
  filter(between(Year, 2005, 2015)) %>%
  ggplot(aes(x = Year, y = Global_Sales)) +
  geom_col() +
  facet_wrap(~Genre) +
  labs(
    title = "Global Video Game Sales (2005-2015) by Genre",
    x = "Year",
    y = "Global Sales (in millions)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_text(margin = margin(t = 10))
  ) +
  scale_y_continuous(labels = scales::comma)

```






## Conclusion

1.  The genre with the best sales is Sports, highlighting its popularity
    among players worldwide.

2.  There is a positive but not particularly strong relationship between
    both user and critic ratings and global sales. Higher scores
    generally align with better sales, but other variables clearly play
    a role. A moderate/strong relationship exists between critic ratings
    and global sales, suggesting their importance as a marker for
    potential bestsellers. This may indicate that critic ratings carry
    more weight or visibility in influencing purchasing decisions.

3.  The most successful platform in terms of global sales is PS2.

4.  X-Box is home to games with the best average ratings, solidifying
    its reputation for quality.

5.  While X-Box has the best average rating, Nintendo has the best sales
    according to this data. The NES and Wii have the highest Q3 of
    Global Sales, with Wii having the most successful game.

6.  We can see the rise of video games in the 2000's with it peaking
    around 2009. From there it gently declines with a sharp decrease
    in 2020. Was this because of data collection and a bias within the
    data, or for actual reasons? Perhaps the pandemic has caused a
    decline in sales recently.

7.  This further shows the popularity of the sports genre, as it is the
    genre with the highest number of sales in all of the genres. Close
    behind, Shooter and Action are also very popular.

8.  While most areas, such as North America, follow the same pattern
    with Action, Shooter, and Sports being the top 3 games, JP follows a
    different trend. JP stands for Japan. Japan's top performing genres
    are Platform, Role-Playing, and Sports. Role-Playing being the
    significantly most popular game genre. This shows the cultural
    difference between the Americas and Japan.

### Further Questions
A few questions we’d like to explore in the future:
First, how do digital-only games compare to physical releases in terms of both sales and ratings? As digital distribution grows, this could reveal important changes in consumer behavior.
Second, what role do major game franchises play in driving platform success?
For example, how much do flagship titles like Call of Duty or Mario influence console popularity?
Third, are strong marketing campaigns or optimal release timing more influential on sales than ratings alone?
This would help us understand if visibility can sometimes outweigh quality in determining success.
And finally, how has the rise of mobile and indie games shifted genre trends over time?
These newer segments may be reshaping the industry in ways that traditional data doesn’t fully capture yet.
These questions could offer deeper insights into how the video game landscape continues to evolve.

