---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Final Project: Video Game Sales

#### Analyn Seeman, Devon Katragadda, Cassidy Berghoff, and Sofi Gutierrez

## Introduction

Intro goes here -- State the goal of the project

1.  Which genre has the best sales?

2.  Is there a relationship between user rating and sales? What about
    critic rating in sales? Which one would be better marker for a
    bestseller?

3.  Which platform had the most successful games?

4.  Which platform's games have the best rating? Does platform affect
    the above questions?

5.  What is the most popular genre per platform?

... and so on

These are the main questions we are looking to answer throughout the
course of the final project. Using the results of the analysis, we will
draw a meaningful conclusion on XXX.

## Data

### Uploading Data Set

```{r}

data = read.csv("Video Game Sales 1978 - 2024 export 2025-03-31 22-10-49.csv")

head(data)
# View(data)
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```


### Description of Data Set

```{r}

str(data)
# View(data)

```

This data set compiles data from video game sales from 1978 to 2024.
There are 16 columns in this data set, representing the 16 variables in
the data set.

### Variables

-   Rank: The index of the data set
-   Name: The name of the game
-   Platform: The platform that the game was released for
-   All_Platforms: A variable that holds a "list" that describes which
    platforms a game is on if it is on multiple
-   All_Games: A list of all of the games in that series
-   Publisher: The publisher of the games
-   Developer: The people that developed the game
-   Critic_Score: What numeric value the critics gave the game (NA,
    0-10)
-   User_Score: What numeric value the users gave the game (NA, 0-10)
-   NA_Sales: Sales in the North American region
-   PAL_Sales: Sales in the European region
-   JP_Sales: Sales in the Eastern Asia region
-   Other_Sales: Sales in all other regions
-   Global_Sales: Year: The year the game was released (if in a series,
    the year the most recent game was released)
-   Genre: The genre of the game

### Cleaning of Data Set

```{r}

q1_data <- data %>%
  drop_na(Global_Sales) #dropping games with no available Sale data

q2_data <- q1_data %>%
  drop_na(Critic_Score) %>%
  drop_na(User_Score) #dropping games with no available Sale data

```

## Marginal Summaries

```{r}

q1_data %>%
  group_by(Platform) %>%
  summarise(maxSales = max(Global_Sales)) #3

# Anything with critic ratings, user ratings, and sales data 
q2_data %>%
  group_by(Platform) %>%
  summarise(maxUserScore = max(User_Score), maxCriticScore = max(Critic_Score)) #4

```

## Results



### Question 1:  Which genre has the best sales?

```{r}
q1_genre <- q1_data %>%
  group_by(Genre) %>%
  summarise(totalSales = sum(Global_Sales)) %>%
  arrange(desc(totalSales))

q1_genre
```


```{r}
ggplot(q1_genre, aes(x = Genre, weight= totalSales)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90))
```

The above table shows the total global sales per video game genre. It is sorted in decreasing order, meaning the most popular genre is located in the top row. The genre with the highest total sales indicates the most popular genre globally. Since `Sports` has the highest `totalSale` value, it is the most popular genre globally.

```{r}
ggplot(data, aes(x = Year, y = log(Global_Sales), color = Critic_Score)) +
  geom_point()
```


### Question 2: Is there a relationship between user rating and sales? What about critic rating?
```{r}

# Scatter plot for user score vs global sales
ggplot(q2_data, aes(x = User_Score, y = Global_Sales)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "User Score vs Global Sales", x = "User Score", y = "Global Sales")

# Scatter plot for critic score vs global sales
ggplot(q2_data, aes(x = Critic_Score, y = Global_Sales)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Critic Score vs Global Sales", x = "Critic Score", y = "Global Sales")

# These scatter plots will visually explore potential correlations between scores (user and critic) and sales. Regression lines provide insights into the strength and direction of the relationships.
```

### Question 3: Which platform had the most successful games?
```{r}
q1_platform <- q1_data %>%
  group_by(Platform) %>%
  summarise(totalSales = sum(Global_Sales)) %>%
  arrange(desc(totalSales))

q1_platform

ggplot(q1_platform, 
       aes(
         x = reorder(Platform, -totalSales), 
         y = totalSales)
       ) +
  geom_bar(stat = "identity") +
  labs(title = "Total Global Sales by Platform",
       x = "Platform",
       y = "Total Global Sales (in millions)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# This will list the platforms ranked by total global sales. The platform with the highest total sales is deemed the most successful.
```

### Question 4: Which platform's games have the best rating?
```{r}
q2_platform <- q2_data %>%
  group_by(Platform) %>%
  summarise(
    avgUserScore = mean(User_Score, na.rm = TRUE),
    avgCriticScore = mean(Critic_Score, na.rm = TRUE)
  ) %>%
  arrange(desc(avgUserScore), desc(avgCriticScore))

q2_platform

# Platforms are ranked based on average user and critic scores. A high average score might indicate quality gaming experiences on that platform
```

### Question 5: What does game success (sales) look like for each platform?

``` {r}
data %>% filter(Platform %in% c("N64", "XB","PS4", "PSP", "PS2", "PS3", "PSN", "NS", "PC", "X360", "DS", "Wii", "NES", "GC", "XOne", "PS", ""), Global_Sales > 5) %>% ggplot(aes(x=Platform, y=Global_Sales)) + geom_boxplot()

```
### Question 6: What years were most successful for video games?
``` {r}

q1_data %>% ggplot(aes(x = Year, y = Global_Sales)) + geom_col()

```

###Question 7: During the best years, which genres were most popular?
``` {r}
q1_data %>% filter(between(Year, 2005, 2015)) %>% ggplot(aes(x = Year, y = Global_Sales)) + geom_col() + facet_wrap(~Genre) + theme(axis.text.x = element_blank(), axis.title.x= element_blank())

```

###Question 8: What is the most popular genre by area?

``` {r}
q1_data %>% ggplot(aes(x = Genre, y = NA_Sales)) + geom_col() + theme(axis.text.x = element_text(size = 8, angle = 45, vjust=0.5))
q1_data %>% ggplot(aes(x = Genre, y = PAL_Sales)) + geom_col() + theme(axis.text.x = element_text(size = 8, angle = 45, vjust=0.5))
q1_data %>% ggplot(aes(x = Genre, y = JP_Sales)) + geom_col() + theme(axis.text.x = element_text(size = 8, angle = 45, vjust=0.5))
q1_data %>% ggplot(aes(x = Genre, y = Other_Sales)) + geom_col() + theme(axis.text.x = element_text(size = 8, angle = 45, vjust=0.5))

```


## Conclusion

1. The genre with the best sales is Sports, highlighting its popularity among players worldwide.

2. There is a positive but not particularly strong relationship between both user and critic ratings and global sales. Higher scores generally align with better sales, but other variables clearly play a role. A moderate/strong relationship exists between critic ratings and global sales, suggesting their importance as a marker for potential bestsellers. This may indicate that critic ratings carry more weight or visibility in influencing purchasing decisions.  

3. The most successful platform in terms of global sales is PS2.

4. XB is home to games with the best average ratings, solidifying its reputation for quality.

5. While XBox has the best average rating, nintendo has the best sales according to this data. The NES and Wii have the highest Q3 of Global Sales, with Wii having the most successful game. 

6. We can see the rise of video games in the 2000's with it peaking around 2009. From there it gently declines with a sharp decrease in 2020. Was this because of data collection and a bias within the data, or for actual reasons? Perhaps the pandemic has caused a decline in sales recently.

7. This further shows the popularity of the sports genre, as it is the genre with the highest number of sales in all of the genres. Close behind, Shooter and Action are also very popular. 

8. While most areas, such as North America, follow the same pattern with Action, Shooter, and Sports being the top 3 games, JP follows a different trend. JP stands for Japan. Japan's top performing genres are Platform, Role-Playing, and Sports. Role-Playing being the significantly most popular game genre. This shows the cultural difference between the Americas and Japan.


